Type-checking Linearity in GHC Core

### 2024-01-23 icfp.tex
.

\documentclass[acmsmall, screen, review]{acmart}
\PassOptionsToPackage{dvipsnames}{xcolor}
\usepackage{boldline}
\usepackage{xcolor}
\usepackage{xargs}
\usepackage{framed}
\usepackage[framemethod=default]{mdframed}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{cmll} % ROMES:TODO: not allowed?
\usepackage{amsthm}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{cleveref}
\usepackage{makecell}
\usepackage{fancyvrb}
\usepackage{thmtools, thm-restate}
\usepackage{listings}
\usepackage{todonotes}
\usepackage[justification=centering]{subfig}

@@

\author{Rodrigo Mesquita}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\affiliation{
   \institution{Well-Typed}
  %\city{City}
   \country{Something}}
 \email{xxx\@yyy.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho\@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

@@

\end{itemize}
%
We review background concepts fundamental to our work in
Section~\ref{sec:background}, including linear type systems, linear types in
Haskell, evaluation strategies, Core's type system, and multiple optimising
transformations applied by GHC in its compilation process.
%
We compare our contributions to related work and discuss possible
avenues for further research (highlighting so-called \emph{multiplicity
coercions}) in Section~\ref{sec:discussion}, which concludes the document.

@@

The next example exposes the case in which the let binder is ignored in the let
body. Here, the linear resource |x| is used in |y|'s body and in the let
body, however, the resource is still used semantically linearly because |y| isn't used at all, thus |x| is consumed just once in the let body:
%
\begin{notyet}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
f2 use x = let y = use x in use x
\end{lstlisting}
\end{notyet}
%
Programmers don't often write bindings that are completely unused, yet, an
optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
also dropped by the optimising compiler} 
from transformations such as inlining, which can substitute out occurrences
of the binder (e.g. |y| is inlined in the let body).

@@

\vspace{-.5cm}
\paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
Second, we look into recursive let bindings. For the most part,
recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
most once} because, when evaluated, the linear resources used in the binders
bodies are consumed. The defining property of a group of mutually recursive let
bindings is that the binders bound in that group can occur, without restrictions, in
the bodies of those same binders. The same way that, in a let body, evaluating
a binding that uses some resource exactly once consumes that resource once,
using the binding in its own definition also entails using that resource once.
Consider the following program, that calls a recursive let-bound function
defined in terms of the linear resource \incode{x} and itself:
%
\begin{notyet}
\begin{lstlisting}
f6 :: Bool -> a *' $\lolli$ '* a
f6 bool x = let go b = case b of
                     True -> x
                     False -> go (not b)
                   in go bool
\end{lstlisting}
\end{notyet}
%
Function \incode{f6} is semantically linear because, iff it is consumed exactly once,
then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
%\begin{itemize}%
%\item
when \incode{bool} is \incode{True}, we'll use the resource \incode{x};
%\item
 when \incode{bool} is \incode{False}, we recurse by calling \incode{go} on \incode{True}, which in turn will use the resource \incode{x}.
%\end{itemize}
In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
other, by recursively calling \incode{go} (which we know will result in using \incode{x}
linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%

@@

There might not be a solution to the set of equations. In this case, the
binding undoubtedly entails using a linear resource more than once. For
example, if we use a linear resource |x| in one case alternative, and invoke
the recursive call more than once, we might eventually consume |x| more than
once:
%
\begin{noway}
\begin{lstlisting}
f9 :: Bool -> Bool *' $\lolli$ '* Bool
f9 bool x =
  let go b
        = case b of
           True -> x
           False -> go (not b) && go True
  in go bool
\end{lstlisting}
\end{noway}
Note that if returned |x| instead of |go bool| in the let body, then, despite
the binding using |x| more than once, we would still consume |x| exactly once,
since recursive bindings are still lazy.

Lastly, we extend our single-binding running example to use two mutually recursive bindings that
depend a linear resource:
\begin{notyet}
\begin{lstlisting}
f10 :: Bool -> a *' $\lolli$ '* a
f10 bool x =
  let go1 b
        = case b of
           True -> go2 b
           False -> go1 (not b)
      go2 b
        = case b of
           True -> x
           False -> go1 b
  in go1 bool
\end{lstlisting}
\end{notyet}
As before, we must find a solution to the set of equations defined by the
mutually recursive bindings to determine which resources will be consumed.
In this case, |go1| and |go2| both consume |x| exactly once if evaluated.
We additionally note that a strongly connected group of recursive bindings
(i.e. all bindings are transitively reachable from any of them) will always
consume the same set of resources -- if all bindings are potentially reachable,
then all linear resources are too.

@@

In short, case expressions enable us to consume resources and thus write
functions that fully consume their linear arguments. To understand exactly how,
we turn to the definition of \emph{consuming} a resource from Linear
Haskell~\cite{cite:linearhaskell}:
\begin{itemize}
    \item To consume a value of atomic base type (such as~\texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate~it.
    \item To consume a function value exactly once, apply it to one argument,
        and consume its result exactly once.
    \item To consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        % For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        % once if pattern-matched on \emph{and} both the first and second element are
        % consumed once.
\end{itemize}
That is, we can consume a linear resource by fully evaluating it, through case
expressions.

@@

In Core, case expressions are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. Critically:
\begin{enumerate}
\item The case scrutinee is always evaluated to Weak Head Normal Form (WHNF).

\item Evaluating to WHNF an expression that is already in WHNF is a no-op, that is,
no computation whatsoever occurs, unlike evaluating a e.g. function application.

\item The case binder is an alias to the result of evaluating the scrutinee to WHNF.

\item The alternative patterns are always exhaustive, i.e. there always exists
a pattern that matches the WHNF of a value resulting from evaluating the
scrutinee, where a pattern is either a wildcard that matches all expressions
($\_$), or a constructor and its linear and non-linear component binders
($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
unrestricted ones).

@@

To explore these case properties in the presence of linearity, we start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{code}
f11 :: a -o b -o (a -o b -o c) -> c
f11 x y use = case (x,y) of { (a,b) -> use a b }
\end{code}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is also
semantically linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{code}
f12 :: a -o b -o (a -o b -o c) -> c
f12 x y use = case (x,y) of z { (a,b) -> use x y }
\end{code}
\end{notyet}
We justify that |f12| is linear by appealing to property 2 -- since the
expression (the tuple) being scrutinized is already in WHNF, evaluating it will
not consume neither $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we aren't using neither $a$ nor $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{code}
f13 :: a -o a -o (a -o a -o c) -> c
f13 x y use = case (x,y) of z { (a,b) -> use a x }
\end{code}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable (or the let binding), we run the computations
defined in terms of the linear variables used for that constructor argument (or
let binder body), but otherwise, if we don't use those binders, then we don't
run the computation thus no resources are consumed.

A third option for this example is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{code}
f14 :: a -o b -o (a -o b -o c) -> c
f14 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{code}
\end{notyet}
Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require doing computation and thus consume linear
resources that are used in it:
\begin{notyet}
\begin{code}
f15 :: a -o b -o (a -o b -o (c,d)) -> (c,d)
f15 x y use = case use x y of z { (a,b) -> z }
\end{code}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we \emph{must} still use either the case binder $z$ or the linear
pattern variables $a,b$, e.g. it would be quite disastrous if any of the following typechecked:
\begin{noway}
\begin{code}
doubleFree :: Ptr -o (Ptr -o Result) -> Result
doubleFree x free = case free x of z { Result v -> free x }
\end{code}
\end{noway}
\begin{noway}
\begin{code}
leakPointer :: Ptr -> ()
leakPointer x = case id x of z { _ -> () }
\end{code}
\end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly to guarantee
that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
|use| in |f15| to simply be |(,)|: it is not sufficient for |use x y| to be
evaluated to WHNF to consume |x| and |y|. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, effectively
discarding linear resources (for cases such as the |use = (,)| example). In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

However, we must also consider pattern matches on constructors without
any linear components. If the constructor has no linear fields, it means the
result can be consumed unrestrictedly and, therefore, all linear resources used
in the computation have been fully consumed.
%
Consequently, in a branch of a constructor without linear fields we know the
result of evaluating the scrutinee to be unrestricted, so we can use the case
binder unrestrictedly and refer to it zero or more times. For example, this
program is semantically linear:
\begin{notyet}
\begin{code}
f16 :: () -o ()
f16 x = case x of z { () -> z <> z }
\end{code}
\end{notyet}
A second example of an unrestricted pattern, where |K2| has no fields and |K1|
has one linear field, seems as though it shouldn't typecheck since the resource
$x$ must have been fully consumed to take the |K2| branch, but because the
scrutinee is known to be |K1|, the |K2| branch is \emph{absurd}, so, in reality
any resource could be freely used in that branch, and the example is
\emph{semantically} linear (despite not being seen as so by our system):
%
\begin{limitation}
\begin{code}
f :: a -o a
f x = case K1 x of z { K2 -> x; K1 a -> x }
\end{code}
\end{limitation}
%
This particular example has a known constructor being scrutinized which might
seem like an unrealistic example, but we recall that during the transformations
programs undergo in an optimising compiler, many programs such as this
naturally occur (e.g. if the definition of a function is inlined in the
scrutinee).
%
% However, the scrutinee before evaluation likely isn't in WHNF, thus scrutinee
% resources cannot directly occur in the case alternatives, since they might be
% potentially consumed when evaluating the scrutinee, the program leading up to
% this one wouldn't have been linear (even semantically).

Further exploring that each linear field must be consumed exactly once, and
that resources in WHNF scrutinees aren't consumed, we are able to construct
more contrived examples, the following two of which the first doesn't typecheck
because the same linear field is used twice, but the second one does since it
uses each linear field exactly once (despite pattern matching on the same
components twice)
\begin{noway}
\begin{code}
f w = case w of z
        (a,b) ->
          case (a,b) of z'
            (c,d) ->
               (a,c)
\end{code}
\end{noway}
\begin{notyet}
\begin{code}
f w = case w of z
        (a,b) ->
          case (a,b) of z'
            (c,d) ->
               (a,d)
\end{code}
\end{notyet}

Before last, we consider the default case alternatives, also known as wildcards
(written $\_$), in the presence of linearity: matching against the wildcard
doesn't provide new information, so linearity is seen as before but without
fully consuming the scrutinee linear resources (in non-linear patterns) nor
binding new linear resources (in linear patterns). In short, if the scrutinee
is in WHNF, we can either use the resources from the scrutinee or the case
binder in that alternative, if the scrutinee is not in WHNF, we \emph{must} use
the case binder, as it's the only way to linearly consume the result of
evaluating the scrutinee.

Finally, we discuss the special case of a case expression scrutinizing a
variable $x$:
\[
\lambda x.~\ccase{x}{\_ \to x}
\]
It might seem as though the program is linear:
\begin{itemize}
\item If $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact.
\item If $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
returning $x$ returns the result of evaluating $x$ that still had to be
consumed exactly once.
\end{itemize}
However, in practice, it depends on the evaluation strategy. If linear function
applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
linear functions use their argument exactly once), and the above function is
considered linear, then an application might duplicate linear resources during
evaluation. For example:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)\\
\Longrightarrow_{CBN~\beta-reduction}\\
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
Therefore, the type system we present in the next section, which evaluates
linear applications call-by-name, does not accept the above program as linear.
Foreshadowing, this particular interaction between evaluation and linearity
comes up in the type preservation proof of our program, and is again explored
with the reverse binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

@@

\subsubsection{\texorpdfstring{$\D$}{Delta}-bound variables}

A $\Delta$-bound variable $u$ is a variable annotated with a usage environment $\Delta$. Crucially, for any $\Delta$-bound variable $u$:
%
\begin{enumerate}
\item Using $u$ is equivalent to using all the linear resources in $\Delta$
%\item $u$ can only be used once, since the $\D$ resources will not be available afterwards
\item Using $u$ is mutually exclusive with using the $\Delta$ resources it depends on elsehow
\item $u$ can be safely discarded as long as the resources in $\Delta$ are consumed elsehow
\end{enumerate}
%
Fortunately, since linear resources must be linearly split across
sub-derivations, (2) follows from (1): consuming the linear resources
in $\Delta$ to type $u$ makes them unavailable in the context of any other
derivation. Therefore, expressions using these resources a second time, directly, or indirectly through the same
(or other) usage environment, is ill-typed, as the resources are
already allocated to the derivation of $u$. Similarly, (3) also follows from
(1), because if the linear resources aren't consumed in the $\Delta$-var
derivation, they must be consumed in an alternative derivation (or otherwise
the expression is ill-typed).

>>>>

\documentclass[acmsmall,review]{acmart}
\PassOptionsToPackage{dvipsnames}{xcolor}
\usepackage{boldline}
\usepackage{xcolor}
\usepackage{xargs}
\usepackage{framed}
\usepackage[framemethod=default]{mdframed}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{cmll} % ROMES:TODO: not allowed?
\usepackage{amsthm}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{cleveref}
\usepackage{makecell}
\usepackage{fancyvrb}
\usepackage{thmtools, thm-restate}
\usepackage{listings}
\usepackage{todonotes}
\usepackage[justification=centering]{subfig}

@@

\author{Rodrigo Mesquita}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\affiliation{
   \institution{Well-Typed}
  %\city{City}
   \country{Something}}
 \email{xxx@yyy.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

@@

\end{itemize}
%
% We review background concepts fundamental to our work in
% Section~\ref{sec:background}, including linear type systems, linear types in
% Haskell, evaluation strategies, Core's type system, and multiple optimising
% transformations applied by GHC in its compilation process.
%
We compare our contributions to related work and discuss possible
avenues for further research
% (highlighting so-called
% \emph{multiplicity coercions})
in Section~\ref{sec:discussion}, which concludes the document.

@@

The next example exposes the case in which the let binder is ignored in the let
body. Here, the linear resource \incode{x} is used in \incode{y}'s body and in the let
body, however, the resource is still used semantically linearly
because \incode{y} isn't used at all, thus \incode{x} is consumed just
once in the let body: 
%
\begin{notyet}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
f2 use x = let y = use x in use x
\end{lstlisting}
\end{notyet}
%
Programmers don't often write bindings that are completely unused, yet, an
optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
also dropped by the optimising compiler} 
from transformations such as inlining, which can substitute out occurrences
of the binder (e.g. \incode{y} is inlined in the let body).

@@

%\vspace{-.5cm}
\paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
Second, we look into recursive let bindings. For the most part,
recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
most once} because, when evaluated, the linear resources used in the binders
bodies are consumed. The defining property of a group of mutually recursive let
bindings is that the binders bound in that group can occur, without restrictions, in
the bodies of those same binders. The same way that, in a let body, evaluating
a binding that uses some resource exactly once consumes that resource once,
using the binding in its own definition also entails using that resource once.
Consider the following program, that calls a recursive let-bound function
defined in terms of the linear resource \incode{x} and itself:
%
\begin{notyet}
\begin{lstlisting}
f6 :: Bool -> a *' $\lolli$ '* a
f6 bool x = let go b = case b of
                     True -> x
                     False -> go (not b)
                   in go bool
\end{lstlisting}
\end{notyet}
%
Function \incode{f6} is semantically linear because, iff it is consumed exactly once,
then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
%\begin{itemize}%
%\item
when \incode{bool} is \incode{True}, we'll use the resource \incode{x};
%\item
 when \incode{bool} is \incode{False}, we recurse by calling \incode{go} on \incode{True}, which in turn will use the resource \incode{x}.
%\end{itemize}
In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
other, by recursively calling \incode{go} (which we know will result in using \incode{x}
linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%

@@

There might not be a solution to the set of equations. In this case, the
binding undoubtedly entails using a linear resource more than once. For
example, if we use a linear resource \incode{x} in one case alternative, and invoke
the recursive call more than once, we might eventually consume \incode{x} more than
once:
%
\begin{noway}
\begin{lstlisting}
f9 :: Bool -> Bool *' $\lolli$ '* Bool
f9 bool x =
  let go b
        = case b of
           True -> x
           False -> go (not b) && go True
  in go bool
\end{lstlisting}
\end{noway}
Note that if returned \incode{x} instead of \incode{go bool} in the let body, then, despite
the binding using \incode{x} more than once, we would still consume \incode{x} exactly once,
since recursive bindings are still lazy.

Lastly, we extend our single-binding running example to use two mutually recursive bindings that
depend a linear resource:
\begin{notyet}
\begin{lstlisting}
f10 :: Bool -> a *' $\lolli$ '* a
f10 bool x =
  let go1 b = case b of
           True -> go2 b
           False -> go1 (not b)
       go2 b = case b of
           True -> x
           False -> go1 b
  in go1 bool
\end{lstlisting}
\end{notyet}
As before, we must find a solution to the set of equations defined by the
mutually recursive bindings to determine which resources will be consumed.
In this case, \incode{go1} and \incode{go2} both consume \incode{x} exactly once if evaluated.
We additionally note that a strongly connected group of recursive bindings
(i.e. all bindings are transitively reachable from any of them) will always
consume the same set of resources -- if all bindings are potentially reachable,
then all linear resources are too.

@@

In short, case expressions enable us to consume resources and so write
functions that fully consume their linear arguments. To understand exactly how,
we turn to the definition of \emph{consuming} a resource from Linear
Haskell~\cite{cite:linearhaskell}:
%begin{itemize}
% \item
to consume a value of atomic base type (such as~\texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate it;
        % \item
        to consume a function value exactly once, apply it to one argument,
        and consume its result exactly once;
        % \item
        To consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        % For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        % once if pattern-matched on \emph{and} both the first and second element are
        % consumed once.
%\end{itemize}
Thus, we can consume a linear resource by fully evaluating it, through case
expressions.

@@

In Core, case expressions are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. Critically,
%\begin{enumerate}
%\item
the case scrutinee is always evaluated to Weak Head Normal Form (WHNF);
%
% \item
evaluating an expression that is already in WHNF is a no-op, that is,
no computation occurs; 
%
the case binder is an alias to the result of evaluating the scrutinee
to WHNF; and,
the alternative patterns are always exhaustive.
% i.e. there always exists
% a pattern that matches the WHNF of a value resulting from evaluating the
% scrutinee, where a pattern is either a wildcard that matches all expressions
% ($\_$), or a constructor and its linear and non-linear component binders
% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
% unrestricted ones).

%\end{enumerate}

@@

To explore these case properties in the presence of linearity, we start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f11 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f11 x y use = case (x,y) of { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is also
semantically linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f12 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f12} is linear by appealing to the fact that the
expression (the tuple) being scrutinized is already in WHNF, evaluating it will
not consume neither $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we aren't using neither $a$ nor $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{lstlisting}
f13 :: a *' $\lolli$ '* a *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f13 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable (or the let binding), we run the computations
defined in terms of the linear variables used for that constructor argument (or
let binder body), but otherwise, if we don't use those binders, then we don't
run the computation thus no resources are consumed.

A third option for this example is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f14 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f14 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require doing computation and thus consume linear
resources that are used in it:
\begin{notyet}
\begin{lstlisting}
f15 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f15 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we \emph{must} still use either the case binder $z$ or the linear
pattern variables $a,b$, e.g. it would be quite disastrous if any of the following typechecked:
\begin{noway}
\begin{lstlisting}
doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
doubleFree x free = case free x of z { Result v -> free x }
\end{lstlisting}
\end{noway}
\begin{noway}
\begin{lstlisting}
leakPointer :: Ptr -> ()
leakPointer x = case id x of z { _ -> () }
\end{lstlisting}
\end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly to guarantee
that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
\incode{use} in \incode{f15} to simply be \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, effectively
discarding linear resources (for cases such as the \incode{use = (,)} example). In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

However, we must also consider pattern matches on constructors without
any linear components. If the constructor has no linear fields, it means the
result can be consumed unrestrictedly and, therefore, all linear resources used
in the computation have been fully consumed.
%
Consequently, in a branch of a constructor without linear fields we know the
result of evaluating the scrutinee to be unrestricted, so we can use the case
binder unrestrictedly and refer to it zero or more times. For example,
the program in Figure~\ref{fig:nolin} is semantically linear.

\begin{figure}[t]
 \begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f16 :: () *' $\lolli$ '* ()
f16 x = case x of z { () -> z <> z }
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{No Linear Fields\label{fig:nolin}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{limitation}
\begin{lstlisting}
f :: a *' $\lolli$ '* a
f x = case K1 x of z { K2 -> x; K1 a -> x }
\end{lstlisting}
\end{limitation}
\vspace{-0.5cm}
\caption{Absurd Branches\label{fig:absurd}}
\end{minipage}
\end{figure}

A second example of an unrestricted pattern, where \incode{K2} has no fields and \incode{K1}
has one linear field, is shown in Figure~\ref{fig:absurd}. Seemingly, the resource
$x$ must have been fully consumed to take the \incode{K2} branch, but because the
scrutinee is known to be \incode{K1}, the \incode{K2} branch is \emph{absurd}, so, in reality
any resource could be freely used in that branch, and the example is
\emph{semantically} linear (despite not being seen as so by our system):
%
%
%
This particular example has a known constructor being scrutinized which might
seem like an unrealistic example, but we recall that during the transformations
programs undergo in an optimising compiler, many programs such as this
naturally occur (e.g.,~if the definition of a function is inlined in the
scrutinee).
%
% However, the scrutinee before evaluation likely isn't in WHNF, thus scrutinee
% resources cannot directly occur in the case alternatives, since they might be
% potentially consumed when evaluating the scrutinee, the program leading up to
% this one wouldn't have been linear (even semantically).

Further exploring that each linear field must be consumed exactly once, and
that resources in WHNF scrutinees aren't consumed, we are able to construct
more contrived examples (Figures~\ref{fig:badcase}
and~\ref{fig:goodcase}). the example in Figure~\ref{fig:badcase} does not typecheck
because the same linear field is used twice, but the example in
Figure~\ref{fig:goodcase} does since it
uses each linear field exactly once, despite pattern matching on the same
components twice.

\begin{figure}[t]
  \begin{minipage}{0.5\textwidth}
\begin{noway}
\begin{lstlisting}
f w = case w of z
        (a,b) -> case (a,b) of z'
                        (c,d) -> (a,c)
\end{lstlisting}
\end{noway}
\vspace{-0.5cm}
\caption{Bad Case\label{fig:badcase}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f w = case w of z
        (a,b) -> case (a,b) of z'
                        (c,d) -> (a,d)
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{Good Case\label{fig:goodcase}}
\end{minipage}
\end{figure}

Finally, we consider the default case alternatives, also known as wildcards
(written $\_$), in the presence of linearity: matching against the wildcard
doesn't provide new information, so linearity is seen as before but without
fully consuming the scrutinee linear resources (in non-linear patterns) nor
binding new linear resources (in linear patterns). In short, if the scrutinee
is in WHNF, we can either use the resources from the scrutinee or the case
binder in that alternative, if the scrutinee is not in WHNF, we \emph{must} use
the case binder, as it's the only way to linearly consume the result of
evaluating the scrutinee. We also discuss the special case of a case
expression scrutinizing a variable $x$:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though the program is linear:
%begin{itemize}
% \item
if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
returning $x$ returns the result of evaluating $x$ that still had to be
consumed exactly once.
%\end{itemize}
However, in practice, it depends on the evaluation strategy. If linear function
applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
linear functions use their argument exactly once), and the above function is
considered linear, then an application might duplicate linear resources during
evaluation. For example:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
Therefore, the type system we present in the next section, which evaluates
linear applications call-by-name, does not accept the above program as linear.
Foreshadowing, this particular interaction between evaluation and linearity
comes up in the type preservation proof of our program, and is again explored
with the reverse binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

@@

\paragraph{\texorpdfstring{$\D$}{Delta}-bound variables}
